Accelerated evolution of convective simulations

### 2018-08-17 tex/bvp_paper.tex
publication-like quality

\newcommand{\Div}[1]{\ensuremath{\nabla\cdot\left( #1\right)}}
\newcommand{\DivU}{\ensuremath{\nabla\cdot\bm{u}}}
\newcommand{\angles}[1]{\ensuremath{\left\langle #1 \right\rangle}}
\newcommand{\KS}[1]{\ensuremath{\text{KS}(#1)}}
\newcommand{\KSstat}[1]{\ensuremath{\overline{\text{KS}(#1)}}}
\newcommand{\grad}{\ensuremath{\nabla}}
\newcommand{\RB}{Rayleigh-B\'{e}nard }
\newcommand{\stressT}{\ensuremath{\bm{\bar{\bar{\Pi}}}}}
\newcommand{\lilstressT}{\ensuremath{\bm{\bar{\bar{\sigma}}}}}
\newcommand{\nrho}{\ensuremath{n_{\rho}}}
\newcommand{\approptoinn}[2]{\mathrel{\vcenter{
	\offinterlineskip\halign{\hfil$##$\cr
	#1\propto\cr\noalign{\kern2pt}#1\sim\cr\noalign{\kern-2pt}}}}}

@@

\begin{abstract}
High Peclet number, turbulent convection is a classic system with a large timescale
separation between flow speeds and the thermal relaxation time.
In this paper, we present a method of
fast-forwarding through the long thermal relaxation of convective simulations, and we test the
validity of this method. This Accelerated Evolution (AE) method involves measuring the dynamics of convection
early in a simulation and using its characteristics to adjust the mean thermodynamic
profile within the domain towards its evolved state. We study \RB convection as a test case for AE.  
Evolved flow properties of AE solutions are measured to be within a few percent
of solutions which are reached through Standard Evolution (SE) over a full thermal diffusion timescale.
At the highest values of the Rayleigh number at which we compare SE and AE,
we find that AE solutions require roughly an order of magnitude fewer cpu-hours
to compute than SE solutions.
\end{abstract}
\maketitle

@@

\section{Introduction}
\label{sec:intro}
Astrophysical convection occurs in the presence of disparate timescales. 
Studying realistic models of natural systems through direct numerical
simulations is infeasible due to the
large separation between various flow timescales and relaxation times.
Stiffness in astrophysical systems can manifest in multiple ways, some of which can
be handled by clever choices of numerical algorithms, and some which cannot.
For example,
flows in the convection zones of stars like the Sun are characteristically low Mach number
(Ma) in the deep interior. Initial value problems solved using
explicit timestepping methods are bound by the Courant-Friedrich-Lewy
(CFL) timestep limit corresponding to the fastest motions (sound
waves), resulting in timesteps which are prohibitively
small for studies of the deep, low-Ma motions. These systems are numerically
stiff, and the difference between
the sound crossing time and the convective overturn time has made studies of low-Ma stellar
convection difficult. This stiffness can be avoided using approximations such as
the anelastic approximation, in which sound waves are explicitly filtered out
\cite{brown&all2010, featherstone&hindman2016}.
Recently, advanced numerical techniques which use fully implicit 
\cite{viallet&all2011, viallet&all2013, viallet&all2016} or mixed
implicit-explicit \cite{lecoanet&all2014, anders&brown2017, bordwell&all2018} 
timestepping mechanisms have made it possible to study
convection in the fully compressible equations at low Mach numbers, 
and careful studies of deep convection which
would have been impossible a decade ago are now widely accessible.

@@

Resolving dynamics in atmospheres which are sufficiently
thermally relaxed therefore remains a challenging problem.
Solar convection is a prime example of this phenomenon, as
dynamical timescales in the solar convective zone are relatively short 
(convection overturns every $\sim 5$ minutes at the solar surface)
compared to the Sun's thermal relaxation timescale, which is O($10^7$) years
\cite{stix2003}.  
In such a system, it is impossible to resolve the convective dynamics while also
meaningfully evolving the thermal structure of the system using
traditional timestepping techniques alone.
As modern simulations aim to model natural, high-Pe convection
in the high-Rayleigh-number (Ra) regime,
the thermal diffusion timescale ($t_{\kappa}$, defined in section \ref{sec:ae}) 
becomes intractably large compared to dynamical timescales 
such as the freefall time ($t_{\text{ff}}$, defined in section \ref{sec:experiment})
\cite{anders&brown2017}, 
\begin{equation}
\frac{t_{\kappa}}{t_{\text{ff}}} \propto (\text{Ra})^{1/2}.
\end{equation}
Furthermore, as dynamical and thermal timescales separate, 
simulations become more turbulent. Capturing appropriately resolved
turbulent motions requires finer grid meshes and smaller timesteps.
Thus, the progression of simulations into the high-Ra
regime of natural convection is slowed by two simultaneous effects: timestepping
through a single convective overturn time becomes more computationally expensive
and the number of overturn times required for systems to thermally relax
grows.

The vast difference between convective and thermal timescales has long plagued
numericists studying convection, and an abundance of approaches has been employed to
study thermally relaxed solutions. One popular method for accelerating the convergence
of high-Ra solutions is by ``bootstrapping'' -- the process of using the relaxed flow
fields and thermal structure of a low Ra solution as initial conditions for a simulation at high
Ra.  This method has been used with great success \cite{johnston&doering2009, verzicco&camussi1997},
but it is not without its faults.  In systems in which there are multiple
stable solutions, such as a roll state and a shearing state of convection,
the large-scale convective structures present in the
low Ra solution imprint onto the dynamics of the new, high Ra solution. 
It is possible that this puts the high Ra solution into a different stable state
than it would naturally reach from an initial, hydrostatically stable configuration.
Another commonly-used tactic in
moderate-Ra simulations is to first solve a simple model of the system in question,
and then use the solution of that model as initial conditions for full nonlinear
direct numerical simulations. For example, past studies have used initial conditions from
a linear eigenvalue solve \cite{hurlburt&all1984} in plane-parallel studies, or
axisymmetric solutions in studies of convection in 3D cylinders \cite{verzicco&camussi1997}. 
In other systems, particularly when convective zones are adjacent to stable regions,
authors often choose initial conditions which are not in a classic, hydrostatic,
conductive state. Rather, either through knowledge of low-Ra solutions \cite{couston&all2017}
or broader convective theories such as Mixing Length Theory
\cite{brandenburg&all2005}, initial conditions can be adjusted such that the stratification
within the convective domain is closer to a relaxed state than the largely
unstable hydrostatic state.

@@

In this work, we study a method of achieving accelerated evolution of
convective simulations. 
Our technique is similar in essence to the approach used in asymptotically 
reduced models of rapidly rotating convection, in which the
mean temperature profile evolves separately from the
fast convective dynamics \cite{julien&all1998, sprague&all2006}. 
We couple measurements of the dynamics of unstable, evolving
convective simulations with knowledge about energy balances in the relaxed solution
to instantaneously and self-consistently adjust the mean vertical thermodynamic 
profile toward its relaxed state. 
While a technique of this kind has been used by many studies previously 
(e.g., \cite{hurlburt&all1986}),
the details of implementation, the convergence properties, and whether or not the
thermally relaxed state achieved from 
accelerated evolution corresponds to the relaxed state in standard evolution 
are not documented.
In section \ref{sec:experiment}, we describe our convective simulations and
numerical methods. In section \ref{sec:ae}, we describe the procedure for 
achieving accelerated evolution. In
section \ref{sec:results}, we compare accelerated evolution solutions
to solutions obtained from standard evolution through a full thermal diffusion timescale. 
In section \ref{sec:speedups}, we compare the numerical cost of accelerated
solutions to standard solutions.
Finally,
in section \ref{sec:extensions}, we offer concluding remarks and
discuss extensions of the methods presented here.

@@

\section{Experiment}
\label{sec:experiment}
In this work we study a simple form of thermal convection:
incompressible \RB convection under the Oberbeck-Boussinesq approximation,
such that the fluid
has a constant kinematic viscosity ($\nu$), thermal diffusivity ($\kappa$), and coefficient
of thermal expansion ($\alpha$). The density of the fluid is a constant, $\rho_0$,
except in the buoyancy term, where it is $\rho = \rho_0(1  - \alpha T_1)$.
The gravitational acceleration, $\bm{g} = - g\bm{\hat{z}}$, is constant.
The equations of motion are \cite{spiegel&veronis1960}:
\begin{gather}
\DivU = 0, 
	\label{eqn:incompressible}
\\
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} =
-\frac{1}{\rho_0}\grad P - g( 1 - \alpha T_1)\hat{z} + \nu\grad^2\bm{u}, 
	\label{eqn:dim_bouss_momentum}
\\
\frac{\partial T_1}{\partial t} + \bm{u}\cdot\grad(T_0 + T_1) = \kappa\grad^2 (T_0 + T_1),
	\label{eqn:dim_bouss_energy}
\end{gather}
where $\bm{u} = u\bm{\hat{x}} + v\bm{\hat{y}} + w\bm{\hat{z}}$ is the velocity, 
$T = T_0(z) + T_1(x, y, z, t)$ are the initial and fluctuating components of temperature, 
and $P$ is the kinematic pressure. The initial temperature profile, $T_0$, decreases
linearly with height.
We non-dimensionalize these equations such that
length is in units of the layer height ($L_z$),
temperature is in units of the initial temperature jump across the layer ($\Delta T_0 = L_z \partial_z T_0$), 
and velocity is in units of the freefall velocity ($v_{\text{ff}} = \sqrt{\alpha g L_z^2 \partial_z T_0}$).
By these choices, one time unit is a freefall time ($t_{\text{ff}} = L_z/v_{\text{ff}}$).
We introduce a reduced kinematic pressure,
$\varpi \equiv (P / \rho_0 + \phi + |\bm{u}|^2 / 2) / v_{\text{ff}}^2$, where the gravitational
potential, $\phi$, is defined such that $\bm{g} = -\grad \phi$. 
In non-dimensional form, and substituting 
$\bm{u}\cdot\grad\bm{u} = \grad(|\bm{u}|^2/2) - \bm{u}\times(\grad\times\bm{u})$
and $\grad^2\bm{u} = -\grad\times(\grad\times\bm{u})$, Eqns.~\ref{eqn:dim_bouss_momentum} \& \ref{eqn:dim_bouss_energy}
become
\begin{align}
\frac{\partial \bm{u}}{\partial t} + \grad \varpi - T_1\hat{z} + \mathcal{R}\grad\times\bm{\omega} &= \bm{u}\times\bm{\omega},
	\label{eqn:bouss_momentum}
\\
\frac{\partial T_1}{\partial t} - \mathcal{P}\grad^2 T_1 + w \frac{\partial T_0}{\partial z} &= - \bm{u}\cdot\grad T_1,
	\label{eqn:bouss_energy}
\end{align}
where $\bm{\omega} = \grad \times \bm{u}$ is the vorticity.
The dimensionless control parameters $\mathcal{R}$ and $\mathcal{P}$ 
are set by the Rayleigh (Ra) and Prandtl (Pr) numbers,
\begin{equation}
\mathcal{R} \equiv \sqrt{\frac{\text{Pr}}{\text{Ra}}}, \qquad \mathcal{P} \equiv \frac{1}{\sqrt{\text{Pr}\,\text{Ra}}}, \qquad
\text{Ra} = \frac{g \alpha L_z^4 \partial_z T_0}{\nu\kappa} = \frac{(L_z\,v_{\text{ff}})^2}{\nu\kappa}, \qquad \text{Pr} = \frac{\nu}{\kappa}.
\end{equation}
We hold Pr $= 1$ constant throughout this work, such that $\mathcal{P} = \mathcal{R}$.
$\mathcal{P}$ and $\mathcal{R}$ are related to the inverse Reynolds and Peclet numbers of the
evolved flows.

In Eqns.~(\ref{eqn:incompressible}), (\ref{eqn:bouss_momentum}), \& (\ref{eqn:bouss_energy}),
linear terms are grouped on the left-hand side of the equations, while nonlinear terms
are found on the right-hand side. We timestep linear terms implicitly, and nonlinear
terms explicitly.
We utilize the 
Dedalus\footnote{\url{http://dedalus-project.org/}} 
pseudospectral framework \cite{burns&all2016} to evolve  
Eqns.~(\ref{eqn:incompressible}), (\ref{eqn:bouss_momentum}), \& (\ref{eqn:bouss_energy}) 
forward in time
using an implicit-explicit (IMEX), third-order, four-stage 
Runge-Kutta timestepping scheme RK443 \cite{ascher&all1997}. The code used to run the simulations
in this work is included in the supplemental materials as a zip file \cite{supp}.

Variables are time-evolved on a dealiased Chebyshev (vertical)
and Fourier (horizontal, periodic) domain in which the
physical grid dimensions are 3/2 the size of the coefficient grid.  
We study 2D and 3D convection in which the domain is a cartesian box, 
whose dimensionless vertical extent is $z \in [0, 1]$, 
and which is horizontally periodic with an extent of $x, y \in [0, \Gamma]$,
where $\Gamma = 2$ is the aspect ratio, as has been previously studied
\cite{goluskin&all2014, johnston&doering2009}. 
In 2D simulations, we set $v = \partial_y = 0$.
We specify no-slip, impenetrable boundary conditions at both the top and
bottom boundary,
\begin{equation}
u = v = w = 0 \, \, \text{at}\,\,z = 0,1.
\label{eqn:vel_bcs}
\end{equation}
The temperature is fixed at the top boundary, and the flux is fixed at the
bottom boundary, such that
\begin{equation}
T_1 = 0 \,\,\text{at}\,\, z=1, \qquad
\frac{\partial T_1}{\partial z} = 0\,\,\text{at}\,\,z=0.
\label{eqn:temp_bcs}
\end{equation}
For this choice of boundary conditions, the critical value of Ra at which
the onset of convection occurs is Ra$_{\text{crit}} = 1295.78$ \cite{goluskin2016}, and the
supercriticality of a run is defined as $S \equiv \text{Ra}/\text{Ra}_{\text{crit}}$.
Studies of convection which aim to model
astrophysical systems such as stars often employ mixed thermal
boundary conditions, as we do here \cite{hurlburt&all1984, cattaneo&all1991, korre&all2017}.
Our choice of the thermal boundary conditions in Eqn.~(\ref{eqn:temp_bcs}) 
was motivated by the fact that accelerated evolution is simpler when both the
thermal profile and the flux through the domain are fixed at a boundary 
(see section \ref{sec:ae}).

@@

\section{The method of Accelerated Evolution}
\label{sec:ae}
Here we describe a method of Accelerated Evolution (AE), which we use 
to rapidly relax the thermodynamic state of convective simulations.  
We compare this AE method to Standard Evolution
(SE), in which we evolve the atmosphere from noise initial conditions
for one thermal diffusion time,
$t_\kappa = \mathcal{P}^{-1}$. Both AE and SE simulations begin with identical
initial conditions, as described in section \ref{sec:experiment}.
As Ra increases, and $\mathcal{P}$ decreases, SE solutions become intractable, 
while the timeframe of convergence for an AE solution remains nearly constant
in simulation freefall time units (see table \ref{table:run_parameters} in
appendix \ref{appendix:run_table}).

We study in depth a 2D simulation at $S = 10^5$ to demonstrate the power of AE.
We compare kinetic energy (KE, black line) and mean temperature (blue line)
traces from a SE run in Fig.~\ref{fig:time_trace}a to an AE run
in Fig.~\ref{fig:time_trace}c.
In Fig.~\ref{fig:time_trace}a, the time evolution of the SE simulation is shown.
The KE grows exponentially from white noise during the
first $\sim 25$ $t_{\text{ff}}$. The solution then saturates and begins to slowly
relax toward the saturated isothermal profile in the interior of the domain.
This slow relaxation is evident in the behavior of the blue line, which measures
$\angles{T} - T_{\text{top}}$, where \angles{T} is the volume-average of $T$, and
$T_{\text{top}} = -0.5$ is the temperature at the upper boundary.
The mean atmospheric temperature and
kinetic energies are fully converged when $t = 4000t_{\text{ff}} = 0.35t_{\kappa}$.
We show roughly the first thousand freefall time
units of evolution, as well as the evolved thermodynamic state reached after a full
thermal time of evolution.  In Fig.~\ref{fig:time_trace}c, similar traces are
shown for the corresponding AE solution
at the same parameters. The same linear growth phase occurs, but shortly after
the peak of convective transient we accelerate the convergence of the atmosphere
through the process which we describe below. We adjust the 1D vertical profile of
the atmosphere three times, as denoted by the three labeled
arrows in the graph numbered 1-3.  The third profile adjustment associated with
arrow 3 is small enough (see appendix \ref{appendix:recipe}) that we assume
the atmosphere is sufficiently converged, and we begin to sample the evolved
convective dynamics.

>>>>

\newcommand{\Div}[1]{\ensuremath{\nabla\cdot\left( #1\right)}}
\newcommand{\DivU}{\ensuremath{\nabla\cdot\bm{u}}}
\newcommand{\angles}[1]{\ensuremath{\left\langle #1 \right\rangle}}
\newcommand{\KS}[1]{\ensuremath{D_{\text{KS}}(#1)}}
\newcommand{\KSstat}[1]{\ensuremath{\overline{D_\text{KS}(#1)}}}
\newcommand{\grad}{\ensuremath{\nabla}}
\newcommand{\RB}{Rayleigh-B\'{e}nard }
\newcommand{\stressT}{\ensuremath{\bm{\bar{\bar{\Pi}}}}}
\newcommand{\lilstressT}{\ensuremath{\bm{\bar{\bar{\sigma}}}}}
\newcommand{\nrho}{\ensuremath{n_{\rho}}}
\newcommand{\approptoinn}[2]{\mathrel{\vcenter{
	\offinterlineskip\halign{\hfil$##$\cr
	#1\propto\cr\noalign{\kern2pt}#1\sim\cr\noalign{\kern-2pt}}}}}

@@

\begin{abstract}
High Peclet number, turbulent convection is a classic system with a large timescale
separation between flow speeds and the thermal relaxation time.
In this paper, we present a method of
fast-forwarding through the long thermal relaxation of convective simulations, and we test the
validity of this method. This Accelerated Evolution (AE) method involves measuring the dynamics of convection
early in a simulation and using its characteristics to adjust the mean thermodynamic
profile within the domain towards its evolved state. We study \RB convection as a test case for AE.  
Evolved flow properties of AE solutions are measured to be within a few percent
of solutions which are reached through Standard Evolution (SE) over a full thermal diffusion timescale.
At the highest values of the Rayleigh number at which we compare SE and AE,
we find that AE solutions require roughly an order of magnitude fewer computing
to evolve than SE solutions.
\end{abstract}
\maketitle

@@

\section{Introduction}
\label{sec:intro}
Astrophysical convection occurs in the presence of disparate timescales. 
Studying realistic models of natural systems through direct numerical
simulations is infeasible because of the
large separation between various flow timescales and relaxation times.
Stiffness in astrophysical systems can manifest in multiple ways, some of which can
be handled by clever choices of numerical algorithms, and some which cannot.
For example,
flows in the convection zones of stars like the Sun are characteristically low Mach number
(Ma) in the deep interior. Initial value problems solved using
explicit timestepping methods are bound by the Courant-Friedrich-Lewy
(CFL) timestep limit corresponding to the fastest motions (sound
waves), resulting in timesteps which are prohibitively
small for studies of the deep, low-Ma motions. These systems are numerically
stiff, and the difference between
the sound crossing time and the convective overturn time has made studies of low-Ma stellar
convection difficult. This stiffness can be avoided using approximations such as
the anelastic approximation, in which sound waves are explicitly filtered out
\cite{brown&all2010, featherstone&hindman2016}.
Recently, advanced numerical techniques which use fully implicit 
\cite{viallet&all2011, viallet&all2013, viallet&all2016} or mixed
implicit-explicit \cite{lecoanet&all2014, anders&brown2017, bordwell&all2018} 
timestepping mechanisms have made it possible to study
convection in the fully compressible equations at low Mach numbers, 
and careful studies of deep convection which
would have been impossible a decade ago are now widely accessible.

@@

Resolving dynamics in atmospheres which are sufficiently
thermally relaxed therefore remains a challenging problem.
Solar convection is a prime example of this phenomenon, as
dynamical timescales in the solar convective zone are relatively short 
(convection overturns every $\sim 5$ min at the solar surface)
compared to the Sun's thermal relaxation timescale, which is O($10^7$) years
\cite{stix2003}.  
In such a system, it is impossible to resolve the convective dynamics while also
meaningfully evolving the thermal structure of the system using
traditional timestepping techniques alone.
As modern simulations aim to model natural, high-Pe convection
in the high-Rayleigh-number (Ra) regime,
the thermal diffusion timescale ($t_{\kappa}$, defined in Sec. \ref{sec:ae}) 
becomes intractably large compared to dynamical timescales 
such as the freefall time ($t_{\text{ff}}$, defined in Sec. \ref{sec:experiment})
\cite{anders&brown2017}, 
\begin{equation}
\frac{t_{\kappa}}{t_{\text{ff}}} \propto (\text{Ra})^{1/2}.
\end{equation}
Furthermore, as dynamical and thermal timescales separate, 
simulations become more turbulent. Capturing appropriately resolved
turbulent motions requires finer grid meshes and smaller timesteps.
Thus, the progression of simulations into the high-Ra
regime of natural convection is slowed by two simultaneous effects: timestepping
through a single convective overturn time becomes more computationally expensive
and the number of overturn times required for systems to thermally relax
grows.

The vast difference between convective and thermal timescales has long plagued
numericists studying convection, and an abundance of approaches has been employed to
study thermally relaxed solutions. One popular method for accelerating the convergence
of high-Ra solutions is by ``bootstrapping'' -- the process of using the relaxed flow
fields and thermal structure of a low Ra solution as initial conditions for a simulation at high
Ra.  This method has been used with great success \cite{johnston&doering2009, verzicco&camussi1997},
but it is not without its faults.  In systems in which there are multiple
stable solutions, such as a roll state and a shearing state of convection,
the large-scale convective structures present in the
low Ra solution imprint onto the dynamics of the new, high Ra solution. 
It is possible that this puts the high Ra solution into a different stable state
than it would naturally reach from an initial, hydrostatically stable configuration.
Another commonly-used tactic in
moderate-Ra simulations is to first solve a simple model of the system in question,
and then use the solution of that model as initial conditions for full nonlinear
direct numerical simulations. For example, past studies have used initial conditions from
a linear eigenvalue solve \cite{hurlburt&all1984} in plane-parallel studies, or
axisymmetric solutions in studies of convection in three-dimensional (3D)
 cylinders \cite{verzicco&camussi1997}. 
In other systems, particularly when convective zones are adjacent to stable regions,
authors often choose initial conditions which are not in a classic, hydrostatic,
conductive state. Rather, either through knowledge of low-Ra solutions \cite{couston&all2017}
or broader convective theories such as Mixing Length Theory
\cite{brandenburg&all2005}, initial conditions can be adjusted such that the stratification
within the convective domain is closer to a relaxed state than the largely
unstable hydrostatic state.

@@

In this work, we study a method of achieving accelerated evolution of
convective simulations. 
Our technique is similar in essence to the approach used in asymptotically 
reduced models of rapidly rotating convection, in which the
mean temperature profile evolves separately from the
fast convective dynamics \cite{julien&all1998, sprague&all2006}. 
We couple measurements of the dynamics of unstable, evolving
convective simulations with knowledge about energy balances in the relaxed solution
to instantaneously and self-consistently adjust the mean vertical thermodynamic 
profile toward its relaxed state. 
While a technique of this kind has been used by many studies previously 
(e.g., \cite{hurlburt&all1986}),
the details of implementation, the convergence properties, and whether or not the
thermally relaxed state achieved from 
accelerated evolution corresponds to the relaxed state in standard evolution 
are not documented.
In Sec. \ref{sec:experiment}, we describe our convective simulations and
numerical methods. In Sec. \ref{sec:ae}, we describe the procedure for 
achieving accelerated evolution. In
Sec. \ref{sec:results}, we compare accelerated evolution solutions
to solutions obtained from standard evolution through a full thermal diffusion timescale. 
In Sec. \ref{sec:speedups}, we compare the numerical cost of accelerated
solutions to standard solutions.
Finally,
in Sec. \ref{sec:extensions}, we offer concluding remarks and
discuss extensions of the methods presented here.

@@

\section{Experiment}
\label{sec:experiment}
In this work we study a simple form of thermal convection:
incompressible \RB convection under the Oberbeck-Boussinesq approximation,
such that the fluid
has a constant kinematic viscosity ($\nu$), thermal diffusivity ($\kappa$), and coefficient
of thermal expansion ($\alpha$). The density of the fluid is a constant, $\rho_0$,
except in the buoyancy term, where it is $\rho = \rho_0(1  - \alpha T_1)$.
The gravitational acceleration, $\bm{g} = - g\bm{\hat{z}}$, is constant.
The equations of motion are \cite{spiegel&veronis1960}:
\begin{gather}
\DivU = 0, 
	\label{eqn:incompressible}
\\
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} =
-\frac{1}{\rho_0}\grad P - g( 1 - \alpha T_1)\hat{z} + \nu\grad^2\bm{u}, 
	\label{eqn:dim_bouss_momentum}
\\
\frac{\partial T_1}{\partial t} + \bm{u}\cdot\grad(T_0 + T_1) = \kappa\grad^2 (T_0 + T_1),
	\label{eqn:dim_bouss_energy}
\end{gather}
where $\bm{u} = u\bm{\hat{x}} + v\bm{\hat{y}} + w\bm{\hat{z}}$ is the velocity, 
$T = T_0(z) + T_1(x, y, z, t)$ are the initial and fluctuating components of temperature, 
and $P$ is the kinematic pressure. The initial temperature profile, $T_0$, decreases
linearly with height.
We non-dimensionalize these equations such that
length is in units of the layer height ($L_z$),
temperature is in units of the initial temperature jump across the layer ($\Delta T_0 = L_z \partial_z T_0$), 
and velocity is in units of the freefall velocity ($v_{\text{ff}} = \sqrt{\alpha g L_z^2 \partial_z T_0}$).
By these choices, one time unit is a freefall time ($t_{\text{ff}} = L_z/v_{\text{ff}}$).
We introduce a reduced kinematic pressure,
$\varpi \equiv (P / \rho_0 + \phi + |\bm{u}|^2 / 2) / v_{\text{ff}}^2$, where the gravitational
potential, $\phi$, is defined such that $\bm{g} = -\grad \phi$. 
In non-dimensional form, and substituting 
$\bm{u}\cdot\grad\bm{u} = \grad(|\bm{u}|^2/2) - \bm{u}\times(\grad\times\bm{u})$
and $\grad^2\bm{u} = -\grad\times(\grad\times\bm{u})$, Eqs.~(\ref{eqn:dim_bouss_momentum}) and (\ref{eqn:dim_bouss_energy})
become
\begin{align}
\frac{\partial \bm{u}}{\partial t} + \grad \varpi - T_1\hat{z} + \mathcal{R}\grad\times\bm{\omega} &= \bm{u}\times\bm{\omega},
	\label{eqn:bouss_momentum}
\\
\frac{\partial T_1}{\partial t} - \mathcal{P}\grad^2 T_1 + w \frac{\partial T_0}{\partial z} &= - \bm{u}\cdot\grad T_1,
	\label{eqn:bouss_energy}
\end{align}
where $\bm{\omega} = \grad \times \bm{u}$ is the vorticity.
The dimensionless control parameters $\mathcal{R}$ and $\mathcal{P}$ 
are set by the Rayleigh (Ra) and Prandtl (Pr) numbers,
\begin{equation}
\mathcal{R} \equiv \sqrt{\frac{\text{Pr}}{\text{Ra}}}, \qquad \mathcal{P} \equiv \frac{1}{\sqrt{\text{Pr}\,\text{Ra}}}, \qquad
\text{Ra} = \frac{g \alpha L_z^4 \partial_z T_0}{\nu\kappa} = \frac{(L_z\,v_{\text{ff}})^2}{\nu\kappa}, \qquad \text{Pr} = \frac{\nu}{\kappa}.
\end{equation}
We hold Pr $= 1$ constant throughout this work, such that $\mathcal{P} = \mathcal{R}$.
$\mathcal{P}$ and $\mathcal{R}$ are related to the inverse Reynolds and Peclet numbers of the
evolved flows.

In Eqs.~(\ref{eqn:incompressible}), (\ref{eqn:bouss_momentum}), and (\ref{eqn:bouss_energy}),
linear terms are grouped on the left-hand side of the equations, while nonlinear terms
are found on the right-hand side. We timestep linear terms implicitly, and nonlinear
terms explicitly.
We utilize the 
Dedalus\footnote{\url{http://dedalus-project.org/}} 
pseudospectral framework \cite{burns&all2016} to evolve  
Eqs.~(\ref{eqn:incompressible}), (\ref{eqn:bouss_momentum}), and (\ref{eqn:bouss_energy}) 
forward in time
using an implicit-explicit (IMEX), third-order, four-stage 
Runge-Kutta timestepping scheme RK443 \cite{ascher&all1997}. The code used to run the simulations
in this work is included in the supplemental materials as a zip file \cite{supp}.

Variables are time-evolved on a dealiased Chebyshev (vertical)
and Fourier (horizontal, periodic) domain in which the
physical grid dimensions are 3/2 the size of the coefficient grid.  
We study two- (2D) and three-dimensional (3D) convection in which the domain is a cartesian box, 
whose dimensionless vertical extent is $z \in [0, 1]$, 
and which is horizontally periodic with an extent of $x, y \in [0, \Gamma]$,
where $\Gamma = 2$ is the aspect ratio, as has been previously studied
\cite{goluskin&all2014, johnston&doering2009}. 
In 2D simulations, we set $v = \partial_y = 0$.
We specify no-slip, impenetrable boundary conditions at both the top and
bottom boundaries,
\begin{equation}
u = v = w = 0 \, \, \text{at}\,\,z = 0,1.
\label{eqn:vel_bcs}
\end{equation}
The temperature is fixed at the top boundary, and the flux is fixed at the
bottom boundary, such that
\begin{equation}
T_1 = 0 \,\,\text{at}\,\, z=1, \qquad
\frac{\partial T_1}{\partial z} = 0\,\,\text{at}\,\,z=0.
\label{eqn:temp_bcs}
\end{equation}
For this choice of boundary conditions, the critical value of Ra at which
the onset of convection occurs is Ra$_{\text{crit}} = 1295.78$ \cite{goluskin2016}, and the
supercriticality of a run is defined as $S \equiv \text{Ra}/\text{Ra}_{\text{crit}}$.
Studies of convection which aim to model
astrophysical systems such as stars often employ mixed thermal
boundary conditions, as we do here \cite{hurlburt&all1984, cattaneo&all1991, korre&all2017}.
Our choice of the thermal boundary conditions in Eqn.~(\ref{eqn:temp_bcs}) 
was motivated by the fact that accelerated evolution is simpler when both the
thermal profile and the flux through the domain are fixed at a boundary 
(see Sec. \ref{sec:ae}).

@@

\section{The method of Accelerated Evolution}
\label{sec:ae}
Here we describe a method of Accelerated Evolution (AE), which we use 
to rapidly relax the thermodynamic state of convective simulations.  
We compare this AE method to Standard Evolution
(SE), in which we evolve the atmosphere from noise initial conditions
for one thermal diffusion time,
$t_\kappa = \mathcal{P}^{-1}$. Both AE and SE simulations begin with identical
initial conditions, as described in Sec. \ref{sec:experiment}.
As Ra increases, and $\mathcal{P}$ decreases, SE solutions become intractable, 
while the timeframe of convergence for an AE solution remains nearly constant
in simulation freefall time units (see table \ref{table:run_parameters} in
appendix \ref{appendix:run_table}).

We study in depth a 2D simulation at $S = 10^5$ to demonstrate the power of AE.
We compare kinetic energy (KE, black line) and mean temperature (blue line)
traces from a SE run in Fig.~\ref{fig:time_trace}(a) to an AE run
in Fig.~\ref{fig:time_trace}(c).
In Fig.~\ref{fig:time_trace}(a), the time evolution of the SE simulation is shown.
The KE grows exponentially from white noise during the
first $\sim 25$ $t_{\text{ff}}$. The solution then saturates and begins to slowly
relax toward the saturated isothermal profile in the interior of the domain.
This slow relaxation is evident in the behavior of the blue line, which measures
$\angles{T} - T_{\text{top}}$, where \angles{T} is the volume-average of $T$, and
$T_{\text{top}} = -0.5$ is the temperature at the upper boundary.
The mean atmospheric temperature and
kinetic energies are fully converged when $t = 4000t_{\text{ff}} = 0.35t_{\kappa}$.
We show roughly the first thousand freefall time
units of evolution, as well as the evolved thermodynamic state reached after a full
thermal time of evolution.  In Fig.~\ref{fig:time_trace}(c), similar traces are
shown for the corresponding AE solution
at the same parameters. The same linear growth phase occurs, but shortly after
the peak of convective transient we accelerate the convergence of the atmosphere
through the process which we describe below. We adjust the 1D vertical profile of
the atmosphere three times, as denoted by the three labeled
arrows in the graph numbered 1-3.  The third profile adjustment associated with
arrow 3 is small enough (see appendix \ref{appendix:recipe}) that we assume
the atmosphere is sufficiently converged, and we begin to sample the evolved
convective dynamics.

The horizontally averaged profiles of the vertical conductive flux, 
F$_{\kappa} = \angles{-\kappa\partial_z T}_{x,y}$, and the vertical convective enthalpy flux,
F$_{\text{E}} = \angles{wT}_{x,y}$, are the basis of the AE method.
Here we use $\angles{}_{x,y}$ to represent a horizontal average. We measure
both of these fluxes early in a simulation, retrieving profiles such as
those shown in Fig.~\ref{fig:time_trace}(b).
As the atmosphere relaxes towards
the isothermal profile specified by the upper (cold) boundary condition, excess
temperature throughout the atmosphere must leave the domain. This excess thermal
energy leaves through the upper boundary, as seen in 
Fig.~\ref{fig:time_trace}(b), where the amount of flux exiting at the top of the domain
is nearly 20 times larger the flux entering the bottom of the domain. Once the atmospheric
temperature profile reaches its evolved state, the flux entering the bottom boundary
is equal to the flux exiting through the upper boundary.  In general, this 
evolution is slow in SE [Fig.~\ref{fig:time_trace}(a)], but AE [Fig.~\ref{fig:time_trace}(c)]
can rapidly advance a system whose fluxes are in a strongly disequilibrium state [Fig.~\ref{fig:time_trace}(b)],
into a near-equilibrium state, as shown in Fig.~\ref{fig:time_trace}(d). In
this final state both boundaries conduct the same amount of flux.
The converged fluxes achieved through AE are at most 5\% different from the SE solution, 
as shown in Fig.~\ref{fig:time_trace}(e). 
This is a very small difference considering the short timescales on which convergence
is reached and the strongly disequilibrium state used to inform the AE process.
